Subset coercions in \Coq

### 2006-08-01 papers/article/main.tex
First real draft of article

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instanciation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence and
of the coercion judgement, which we don't need here as we have unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respect some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again this wasn't adapted to our setting, so we invented this
new presentation.

@@

Next, we have the rules for products and sums, which compose
coercions. In the \irule{SubProd} rule, we first coerce the argument to
an object of type $\ip{T}{`G}$. Then we apply it to the coerced
function, resulting in an object $\ctxdot~(c_1[x])$ of type
$\ip{V}{`G, x : T}[c_1[x]/x]$. We'll see later that this type is equivalent to
$\ip{V}{`G, x : U}$, so we can apply the second coercion and get an
object of type $\ip{W}{`G}$. Similarly for dependent sums \TODO{}

Now for the original part, the \irule{SubSub} rule simply inserts a
projection before applying the inductively defined coercion on the
support of the subset. On the other hand, \irule{SubProof} creates and
object of a subset using its sole constructor \sref{elt}. The object
part is given by the coercion between the support while the proof part
is promised using a metavariable of type $\ip{P}{`G, x : U}[c/x]$ which
we will turn later into a proof obligation.

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms to a system with only
$\beta\pi$-equivalence. Indeed, if the algorithmic coercion judgement is transitive
and has a unicity property, then we may suppose that the coercions
themselves have corresponding properties:
\begin{proposition}[Coercion transitivity]
  If $\subimpl{`G}{c_1}{T}{U}{s}$ and $\subimpl{`G}{c_2}{U}{V}{s}$ then
  there exists a coercion $c$ so that $\subimpl{`G}{c}{T}{V}{s}$ and $c
  \eqqbp c_2 `o c_1$ (where $c_2 `o c_1 = c_2[c_1]$).
\end{proposition}

Now take the following example
\begin{example}[Transitivity needs eta]
  Let $T, U, V = \Pi x : A.B$.
  Then $c_1$ (and $c_2$ by unicity) must have the form $\lambda x :
  \ip{C}{`G}.c^2_1[\ctxdot~c^1_1[x]]$ (respectively $\lambda x : \ip{A}{`G}.c^2_2[\ctxdot~c^1_2[x]]$).
  So $c_2 `o c_1 = \lambda x : \ip{A}{`G}. c^2_1[(\lambda x :
  \ip{A}{`G}.c^2_2[\ctxdot~c^1_2[x]])~c^1_1[]] \redbp \lambda x :
  \ip{A}{`G}. c^2_1[c^2_2[\ctxdot~c^1_2[c^1_1[x]]]]$.
  Suppose all $c^i_j$ have been derived by \irule{SubConv}. Then we have 
  $c_2 `o c_1 = $.

\end{example}

\ccieqfig[Equational theory of \CCq]

\subsection{Properties}
- Proofs.

\section{Conclusion}
\subsection{Related work}

\subsection{Further work}

\newpage

>>>>

Coercions are formalized as multi-holes evaluation contexts, with
$\ctxdot$ the denotation of a hole. We define the instanciation $C[x]$ of a context $C$
by a term $x$ as simultaneous substitution of $x$ for every $\ctxdot$ in $C$.
Hence if $C$ is an evaluation context and $x$ a term, $C[x]$ is a term
(ie, it cannot contain $\ctxdot$).
The holes will denote the object to which the coercion is applied. Other
presentations are possible, notably the ``\LMS{}'' style of \cite{longo95logic} which
was adopted by Gang Chen in his treatment of coercive subtyping for the
Calculus of Constructions \cite{conf/popl/Chen03}, and the presentation 
of Amokrane Sa\"ibi \cite{saibi97inheritance}. The \LMS{} style consists
in defining a judgement $x : A \type y : B$ which reads: $A$ is
coercible to $B$ and the coercion is $\lambda x : A.y$. This
presentation is very elegant and well suited for checking coherence and
of the coercion judgement, which we don't need here as we can get unicity
of coercions. Moreover, it did not seem readily usable in
our setting, where coercions may contain metavariables. Similarly, Amokrane Sa\"ibi defines
a coercion derivation algorithm which respect some coherence criteria,
roughly, that type-erased coercions $\beta\eta$-reduce to the
identity. Again, this wasn't adapted to our setting, so we invented this
new presentation.

@@

Next, we have the rules for products and sums, which compose
coercions. In the \irule{SubProd} rule, we first coerce the argument to
an object of type $\ip{T}{`G}$. Then we apply it to the coerced
function, resulting in an object $\ctxdot~(c_1[x])$ of type
$\ip{V}{`G, x : T}[c_1[x]/x]$. We'll see later that this type is equivalent to
$\ip{V}{`G, x : U}$, so we can apply the second coercion and get an
object of type $\ip{W}{`G}$. There is a little twist when coercing the
second component of a dependent sum. Indeed, we must simultaneously
substitute the first projection of the coerced sum in the second
coercion which was typed in environment $`G, x : T$ and also instanciate
the coercion with the second component. Clearly, the $\ctxdot$ of the
former shouldn't be instanciated with the later, so we must first
instanciate $c_2$ and then substitute $\pi_1~\ctxdot$ for $x$, knowing
the $x$ can't appear in $\pi_2~t$ where $t : \Sigma x : T.V$.

Now for the original part, the \irule{SubSub} rule simply inserts a
projection before applying the inductively defined coercion on the
support of the subset. On the other hand, \irule{SubProof} creates an
object of a subset using its sole constructor \sref{elt}. The object
part is given by the coercion to the support of the target subset
while the proof part is promised using a metavariable of type $\ip{P}{`G, x : U}[c/x]$ which
we will turn later into a proof obligation.

@@

\subsection{A little more expressiveness}
One may wonder how can we interpret \Russell terms to a system with only
$\beta\pi$-equivalence. 
Indeed, if our algorithmic system has transitivity and symetry of
coercions, we expect that the derived coercions will inherit these
properties, for example:
\begin{proposition}[Symetry of coercions]
  If $\subimpl{`G}{c}{T}{U}{s}$ then there exists $c^{-1}$ so that
  $\subimpl{`G}{c^{-1}}{U}{T}{s}$ and $c `o c^{-1} `= c^{-1} `o c `=
  \ctxdot$.
\end{proposition}

Now take the following example:
\begin{example}[Symetry for products]
  Let $T, U = \Pi x : A.B$. The coercions $c, c^{-1}$ derived by
  the algorithm must have the form $\lambda x :
  \ip{A}{`G}.c_2[\ctxdot~c_1[x]]$, hence we must include the $\eta$
  rule in order to have $c `o c^{-1} `= \ctxdot$.
\end{example}

Similarly, we must add $\eta$ rules for all introduction terms of our
language if we want to be conservative (the coercion algorithm will
always derive $\eta$-long normal forms, so we must be able to prove that they
are equivalent to the empty coercion) and have unicity of
coercions. The later is an important property of our system, as it
corresponds in practice to the determinism of proof-obligations
generation. The complete equational theory of \CCq{} is given in figure
\vref{fig:eqccq}. It includes $\eta$, surjective pairing for dependent
sums and subsets ($\rho$) and proof-irrelevance for the second component
of subset objects.
Benjamin Werner has studied the addition of proof-irrelevance (rule
$\sigma$) in the calculus of constructions
\cite{Werner:ProofIrrelevance}. It is not a trivial extension and it has
far-reaching consequences on the model of the calculus
\cite{DBLP:conf/types/MiquelW02}. We direct the reader to these papers for further
information on the subject.

\ccqeqfig[Equational theory of \CCq]

\subsection{Properties}
The correctness proof of the traduction is very involved, so we will
only sketch it here. A paper version \vref{sozeau:coq/Russell/report} is
available (in french) and a mechanically checked proof is in the works.

We first prove reflexivity, symmetry and transitivity of the coercion
derivation algorithm, which extends previous proofs for the algorithmic
system with properties on
the generated coercions. Then we show substitutivity of the
interpretation: if $`G, x : U, `D \type t : T$ then $\ip{t[u/x]}{`G,
 `D[u/x]} = \ip{t}{`G, x : U, `D}[\ip{u}{`G}/x]$. We extend the notion
of coercion to contexts next and show stability of judgements under
context coercion. It is then possible to prove a lemma about commutativity of
subtitution with the interpretation and coercions. Then we get an
important corrolary: if $`G \type u : U$, $\subimpl{`G}{c}{U}{V}{s}$ and
$`G, x : V \type T : s$ then $\ip{T[u/x]}{`G} `= \ip{T}{`G, x :
  V}[c[\ip{u}{`G}]/x]$. Essentially, it shows that coercions inserted
by the interpretation depend only on the context and hence could be
added later using substitution instead. This allows us to show that
equivalence is conserved by interpretation and finaly that our
interpretation is correct.

\section{Conclusion}
We have developed a new language for writing programs in the \Coq proof
assistant which allows the user to specify complex programs while
keeping the corresponding code simple. It is a first step towards
interpretation of \ML code into \Coq for building certified
programs. While the system and its accompanying proofs may seem complex, the
implementation is short, simple and has been tested with success on
various examples. This works is also a second proof of the
meaningfullness of the Predicate Subtyping feature of \PVS{} and may
help build a common interface for the two provers.

\subsection{Related work}
Other efforts to build certified programs using type theory include
Cayenne \cite{Augustsson99} and Epigram
\cite{DBLP:journals/jfp/McBrideM04}. Cayenne offers dependent
types and general recursion at the expense of a non-terminating type
checker, it can be sought of as a testbed for developing languages with
dependent types but is not so much aimed at building certified programs.
Epigram ought to become a complete programming language with dependent
types at its heart. Instead of using a phase distinction to separate
coding and proving, Epigram is based on an interactive, 2-dimensional editing process
where code and proofs are written incrementaly to get a complete program.
Using type annotations and with the help of the editor for structuring, 
they are able to write programs with precise specifications. However it
is not obvious if it could be made usable by programmers because of the
paradigm change and scalling issues.

\subsection{Further work}
Much has to be done for integrating \Russell cleanly into \Coq. The
treatment of existential variables of \Coq has to be improved, the
type inference algorithm needs some tuning to become more similar to \ML
and an implementation of proof-irrelevance, while rarely needed in
practice, is necessary to have a robust system.
We also hope to extend this mechanism of proof-obligations generation to other
constructs, notably inductive types. Finally, we intend to use this system as a
basis for interpretation of \ML{} programs in \Coq, using a monadic
traduction or some kind of effects as first class objects.
